---
title: "Data Science Wages by Industry Group and State"
subtitle: "An Analysis of Bureau of Labor Statistics Survey Data, May 2022"
author: "Lyndsay Miles"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(readr)
library(kableExtra)
library(ggplot2)
library(reactable)
library(DT)
library(plotly)
library(readxl)
```

```{r, echo=FALSE, include=FALSE}
#coli data from kaggle
coli <- read.csv("data/cost_of_living_missouri_economic_research_and_information_center.csv") 
salary <- read.csv("data/oes_research_2022_allsectors_trimmed.csv")
#salary <- read_excel("data/oes_research_2022_allsectors_abbr.xlsx", sheet=1)
#states data from https://developers.google.com/public-data/docs/canonical/states_csv
states <- read.csv("data/states.csv")
```

```{r, echo=FALSE, include=FALSE}
salary_clean <- salary %>% 
  rename_with( ~ tolower(gsub(" ", "", .x, fixed = TRUE))) %>%
  select(area, area_title, naics, naics_title, occ_title, tot_emp, a_mean) %>%
  rename(name = area_title)
```

```{r, echo=FALSE, include=FALSE}

salary_clean$a_mean <- as.numeric(salary_clean$a_mean)
salary_clean$tot_emp <- as.integer(salary_clean$tot_emp)
#for expedience, I decided to drop NAs for now. 
salary_clean <- salary_clean %>% drop_na()

```

```{r, echo=FALSE, include=FALSE}
coli_clean <- coli %>% 
  rename_with( ~ tolower(gsub(" ", "", .x, fixed = TRUE))) %>%
  select(state, conversion) %>%
  mutate(conversion_adj = conversion/100)

#remove non-US states
coli_clean <- coli_clean[!(coli_clean$state == "British Columbia"|
                             coli_clean$state == "Ontario"),]
  
```

```{r, echo=FALSE, include=FALSE}
#join the 3 datasets 
salary_joined <- salary_clean %>%
  left_join(states, by = "name")%>%
  left_join(coli_clean, by = "state") 
```

```{r, echo=FALSE, include=FALSE}
#add an adjusted salary, weighted by the cost of living index

salary_joined_adj <- salary_joined %>% 
  group_by(name, naics, naics_title, occ_title) %>%
  mutate(adjusted_sal = (a_mean * (1+(1-(conversion/100) )))) %>%
  arrange(name, naics, naics_title, occ_title)
```

```{r, echo=FALSE, include=FALSE}
#just data scientists
salary_ds <- salary_joined_adj %>% 
  filter(occ_title == "Data Scientists")
```

```{r, echo=FALSE}
#barchart 

ds_jobs_num_table <- salary_ds %>%
  group_by(naics_title) %>% 
  summarize(num_jobs = sum(tot_emp)) %>% 
  arrange(desc(num_jobs))
ds_jobs_num_table %>% 
  
ggplot(aes(x = reorder(naics_title, num_jobs), y = num_jobs, fill=naics_title)) + 
  geom_col() +
  scale_x_discrete(label=function(x) stringr::str_trunc(x, 20))+
  labs(x = "", y = "", title = "Figure 1: Total number of persons employed as data scientists", subtitle = "May 2022, by NAICS group")+
  theme(axis.text.x = element_text(size = 8, angle = 45, hjust = 1, vjust = 0.9),
        legend.position="none")

```


```{r, echo=FALSE}
#Table of epi salaries by NAICS group with adjusted salaries
ds_jobs_groups_table <- salary_ds %>%
  group_by(naics_title) %>%
  summarize(mean_adj_income = mean(adjusted_sal),
            num_jobs = sum(tot_emp)) %>% 
  arrange(desc(mean_adj_income))
  
kable(ds_jobs_groups_table,
      booktabs = T,
      col.names = c("Group title", "Mean adjusted annual wage ($US)", "Total sampled"),
      caption = "Table 1: Mean adjusted income by group for data scientists (adjusted by cost of living)") %>%
  kable_styling(full_width = F)
```

```{r, include=FALSE, echo=FALSE}


ds_jobs_num_table <- data.frame(ds_jobs_num_table) 
ds_jobs_num_table$naics_title <- as.factor(ds_jobs_num_table$naics_title)

kable(ds_jobs_num_table,
      booktabs = T,
      col.names = c("Group title", "Number of employed"),
      caption = "Number of data scientists employed by NAICS group") %>%
  kable_styling(full_width = F)
```

```{r, echo=FALSE, include=FALSE}
#what are number of jobs available by state?
#in what state do you have the most jobs and also the most pay for data scientists?
#for example, New Jersey is fairly high on the list in terms of income and also high # of jobs (11 out of 54 - includes Puerto Rico, Guam, US Virgin Islands, and Washington DC)
#assign each state a # between 1 and 54 for their income level & another assign # for their job availability
ds_jobs_by_state <- salary_ds %>% 
  group_by(name) %>% 
  summarize(tot_jobs = sum(tot_emp),
            avg_pay = round(mean(a_mean),2)) %>% 
  mutate(rank_most_jobs = percent_rank(tot_jobs),
         most_jobs = ifelse(rank_most_jobs >= 0.75, 1, 0),
         rank_most_pay = (percent_rank(avg_pay)),
         most_pay = ifelse(rank_most_pay >= 0.75, 1,0),
         best_both = ifelse(most_jobs == 1 & most_pay == 1, 1, 0)
         ) %>%
  arrange(desc(best_both), desc(most_pay))

ds_jobs_by_state_table <- ds_jobs_by_state %>%
  select(name, tot_jobs, avg_pay, rank_most_jobs, rank_most_pay, best_both)
```

```{r, echo=FALSE, message=FALSE}
datatable(
  data = ds_jobs_by_state_table,
  class = 'cell-border stripe',
  rownames = FALSE,
  colnames = c("U.S. State", "Total data scientists sampled", "Annual adjusted mean wage ($US)", "State rank for # of data scientists sampled (1=highest)", "State rank for wage value (1=highest)", "States ranked for both rankings (1=both)"),
  extensions = 'Buttons', options = list(dom = 'Bfrtip',
                                                 buttons = c('copy', 
                                                             'csv', 
                                                             'excel', 
                                                             'pdf', 
                                                             'print'),
                                                 #how many entries per page
                                                 pageLength = 10,
                                                 #options for "Show how many entries"
                                                 lengthMenu = c(10, 20, 30))) %>%
  formatStyle("name", fontWeight = "bold", color = "#2A0D5D") %>%
  formatRound("avg_pay", 2) %>%
  formatRound("rank_most_jobs", 2)%>%
  formatRound("rank_most_pay", 2)%>%
  formatStyle("best_both", 
              target = "row",
              backgroundColor = styleEqual(c(0,1), c("white", "#ADDFB3")))
```



```{r, echo=FALSE, message=FALSE}
plot1 <- ds_jobs_by_state %>% 
  ggplot(aes(
    x=tot_jobs, 
    y=avg_pay, 
    text = paste("\nName:", name))) +
  geom_point()+
  #geom_smooth(method="lm", formula= "y~x")+
  geom_smooth(span=0.3)+
  scale_x_log10()+
  labs(x="number employed on log10 scale", y="Average adjusted annual wage ($US)", caption = "Data source: U.S. Bureau of Labor Statistics")+
  theme_bw()
#plot1
ggplotly(plot1)
```

```{r, echo=FALSE, include=FALSE}
#calculate intercept/slope
coef1 <- coef(lm(avg_pay~tot_jobs, data=ds_jobs_by_state))
#9.732738e+04
#3.231054e-01 
cor(ds_jobs_by_state$tot_jobs, ds_jobs_by_state$avg_pay)
#0.0.2622449
```
### Data

I used the most recent data tables provided by the U.S. Bureau of Labor Statistics (Research estimates by state and industry/all sectors). This data is derived from sample surveys, which are subject to sampling and non-sampling error. To take into consideration variations in cost of living from state to state, I adjusted salary values using the Missouri Economic Research and Information Center (MERIC) cost of living estimates for 2023. According to their website, MERIC arrives an the cost of living index by averaging the indices of cities and metropolitan areas in a state.

To conduct this analysis, I cleaned and joined these two data sets and filtered to only include observations where the occupational title is "Data Scientists". I used the following variables:

-   State name

-   North American Industry Classification System (NAICS) title for the given industry

-   Standard Occupational Classification title or OEWS specific title for the occupation

-   Mean annual wage

-   Estimated total employed rounded to the nearest 10 (excludes self-employed).

I then calculated the mean adjusted annual wage by creating a new variable using this formula:

> $adjusted salary = (mean Annual Salary(1 + 1-cost Of Living Index/100))$

### Conclusions

-   The largest portion of data scientists employed and sampled in the BLS survey as employed in May 2022 (107,840) worked in professional, scientific, and technical services.  

-   Those with the largest incomes work in, e.g., Merchant Wholesalers, Nondurable Goods (4242 and 4246 only), Semiconductor and Other Electronic Component Manufacturing, and Web Search Portals, Libraries, Archives and Other Information Services. The mean adjusted annual wages for those employed in these sectors is $133,075 to $140,875. 

-   Four states score highly in terms of employing high numbers of data scientists and paying the high average salaries: California, Massachusetts, New York, and Virginia. 

-   Hawaii and Montana employed the fewest data scientists Annual adjusted mean wages for data scientists were lowest in Montana and Mississippi.

### Limitations

-   The salary data used is from May 2022, and more recent data would be beneficial.

-   For brevity, I chose to focus on mean adjusted annual income, but looking at wage ranges would be meaningful as well.

-   There are many other job titles that data scientists work under. This analysis only captures wages and number employed and surveyed with the title "data scientists".

-   This does not include data from all 50 states; only observations from 46 states were available in the survey.

### References

U.S. Bureau of Labor Statistics. (2023, April 27). Tables created by BLS. U.S. Bureau of Labor Statistics. Retrieved March 21, 2024 from <https://www.bls.gov/oes/tables.htm>.

Karol Łukaszczyk. (June 2023). Cost of living in State in USA - MERIC (adapted from Missouri Economic Research and Information Center's Cost of Living Data Series), Version 1. Retrieved March 21, 2024 from <https://www.kaggle.com/datasets/lukkardata/cost-of-living-missouri-economic-research>.

Link to the code for this analysis: <https://github.com/LyndsayM/salaries>
